What do they see? Testing perceptual accuracy in a U.S. general population survey using stacked bar charts

### 2023-10-25 SDSS-2023/testing-charts-accuracy.qmd
title change

---
title: "What do they see? Testing perceptual accuracy  in surveys of the of the general US population using stacked barcharts"
short-title: Testing stacked barcharts in surveys # running title
bibliography: references.bib
abstract: "The use of visuals is a key component in scientific communication. Decisions about the design of a data visualization should be informed by what design elements best support the audience's ability to perceive and understand the components of the data visualization. We build on the foundations of Cleveland and McGill's work in graphical perception, employing a large, nationally-representative, probability-based panel of survey respondents to test perception in stacked bar charts. Our findings provide actionable guidance for data visualization practitioners to employ in their work."
keywords: 'graphical perception, data visualization, survey data'
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
  \usepackage{booktabs}
  \usepackage{graphicx}
  \usepackage{hyperref}
  \usepackage{setspace}
  \usepackage{lscape}
documentclass: jds
format:
  pdf: 
    includes:
      in_header: 'gt_packages.sty'
    include-in-header: 
      text: |
        \usepackage{lscape}
        \author[1]{Kiegan Rice\thanks{Corresponding author. Email: rice-kiegan@norc.org}}
        \author[2]{Heike Hofmann\footnote{Email: hofmann@iastate.edu}}
        \author[1]{Nola du Toit\footnote{Email: dutoit-nola@norc.org}}
        \author[1]{Edward Mulrow\footnote{Email: mulrow-edward@norc.org}}
        \affil[1]{NORC at the University of Chicago}
        \affil[2]{Department of Statistics, Iowa State University}
    keep-tex: true
editor: 
  markdown: 
    wrap: sentence
---

@@

# To-do

Into appendix: Stimulus question(s)

@@

```{r}
#| label: fig-tasks
#| fig-cap: "Overview of all of the visual stimuli presented to viewers. In each bar chart, two pieces are marked. The larger piece in each chart are pieces B and D. Please refer to the appendix for full size images of the stimuli."
#| out-width: 100%
knitr::include_graphics("images/stimuli-overview-1.png")
```

@@

| Name      | Date       |      \# Participants |                Effective Sample Size |         Sum of weights $\sum_i w_i$ |                                  $\lambda_i$ |
|------------|------------|-----------:|-----------:|-----------:|-----------:|
| Round 1   | April 2022 | `r nrow(round1_adj)` | `r round(neff(round1_adj$weight),1)` | `r round(sum(round1_adj$weight),1)` |        $\lambda_1$ = `r round(lambdas[1],3)` |
| Round 2   | May 2022   | `r nrow(round2_adj)` | `r round(neff(round2_adj$weight),1)` | `r round(sum(round2_adj$weight),1)` | $\lambda_2$ = `r sprintf("%.3f",lambdas[2])` |
| Round 3   | June 2022  | `r nrow(round3_adj)` | `r round(neff(round3_adj$weight),1)` | `r round(sum(round3_adj$weight),1)` |        $\lambda_3$ = `r round(lambdas[3],3)` |
| **Combined** | ---        |  `r nrow(round_123)` |  `r round(neff(round_123$weight),1)`* |                                     |                                              |

@@

```{r}
#| label: fig-response-123
#| fig-cap: "Responses by task across the designs. The overlaid rectangles represent 95% confidence intervals. The letters in blue and orange encode significances between pairwise proportions [@piephoAlgorithmLetterBasedRepresentation2004]: two bars have a significantly different proportion (at a 5% significance level) if they do not have a letter in common. There is no significant difference between the designs for wrong responses."
#| fig-height: 4
#| fig-width: 8
#| out-width: 90%

@@

where $\ell \in \{1, 2\}$; $i \in \{1, 2\}$ is the comparison type (1 = Aligned, 2 = Unaligned), and $j \in \{1, 2, 3\}$ is the chart design, with 1 = Vertical, 2 = Horizontal, and 3 = Horizontal wide.
The letters allow us to assess whether differences between two means are significant:  two estimates within the same column are significantly different (at the 5% level) if they do not have any letter in common. This method is called connected letters and has been introduced by @piephoAlgorithmLetterBasedRepresentation2004. In @fig-response-123, we see, that the proportions of wrong responses are significantly different between tasks, but designs within each task are not significantly different. For correct responses, all proportions are significantly different except for the proportion of correctly assessing the difference between unaligned rectangles in the wide horizontal design: their label of `ab` shows us that assessing the differences between unaligned rectangles in the wide horizontal design is not significantly better than in the horizontal design (labelled `a`), but also not significantly worse than in the vertical design (labelled `b`).

The estimated values and 95% confidence intervals are shown in @tbl-rounds-123.

@@

```{r}
#| label: fig-zoom-123
#| fig-cap: "Rates of zooming by task, chart design, and correctness. The overlaid rectangles represent 95% confidence intervals. There are significantly higher rates of zooming by respondents for the vertical task compared to both horizontal tasks; rates of zooming do not differ by correctness of response or alignment in the stimulus."
#| fig-height: 3.5
#| fig-width: 11
#| out-width: 90%

confint_zoom1 <- svyby(formula = ~zoom, by = ~task+response3, design = survey_123, svyciprop) %>%
  mutate(task = factor(task, levels=c("cd", "ab"), labels=c("Aligned", "Unaligned")), 
         zoom2 = zoom, 
        response3 = factor(response3, levels=rev(c("Correct", "They are the same", "Wrong"))),
        response3 = fct_recode(response3, "They are\nthe same" = "They are the same")) %>%
  dplyr::select(-c("zoom"))

confint_zoom2 <- svyby(formula = ~zoom, by = ~chart+response3, design = survey_123, svyciprop) %>%
  mutate(chart = factor(chart, levels=c("Vertical", "Horizontal", "Horizontal wide")), 
         zoom2 = zoom, 
        response3 = factor(response3, 
                           levels=rev(c("Correct", "They are the same", "Wrong"))),
        response3 = fct_recode(response3, "They are\nthe same" = "They are the same")) %>%
  dplyr::select(-c("zoom"))

zt <- round_123_long %>% 
  mutate(
    response3 = c("Wrong", "They are the same", "Correct")[correct+correct3+1],
    response3 = factor(response3, levels=rev(c("Correct", "They are the same", "Wrong"))),
    response3 = fct_recode(response3, "They are\nthe same" = "They are the same"),
    task = factor(task, levels=c("cd", "ab"), labels=c("Aligned", "Unaligned")),
    chart = factor(chart, levels=c("Vertical", "Horizontal", "Horizontal wide")),
    zoom = factor(zoom, levels=c("no", "yes"))
  ) %>% 
  ggplot(aes(x = response3)) + 
  geom_bar(aes(fill = zoom, weight = weight), position="fill") +
  facet_grid(.~task) + 
  theme(axis.text.x = element_text(angle=0, vjust = 0.5)) + 
  scale_fill_manual("Zoom", values=c("grey", "#1274F3")) +
  xlab("") + 
  ggtitle("Zooming behavior\nby task") +
  scale_y_continuous("Percentage", breaks=seq(0,1, by=0.25), 
                     labels = seq(0,1, by=0.25)*100) + 
  geom_rect(
    aes(
      xmin = as.numeric(response3) - 0.45,
      ymin = zoom2 - qnorm(0.975)* `se.as.numeric(zoom)`,
      xmax = as.numeric(response3) + 0.45,
      ymax = zoom2 + qnorm(0.975)* `se.as.numeric(zoom)`,
    ), 
    fill = "black", alpha = 0.25, #colour = "white", linewidth = 0.25,
    data = confint_zoom1
  )  + 
  geom_segment(
    aes(
      x = as.numeric(response3) - 0.45,
      y = zoom2,
      xend = as.numeric(response3) + 0.45,
      yend = zoom2
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = confint_zoom1
  )

zd <- round_123_long %>% 
  mutate(
    response3 = c("Wrong", "They are the same", "Correct")[correct+correct3+1],
    response3 = factor(response3, levels=rev(c("Correct", "They are the same", "Wrong"))),
    response3 = fct_recode(response3, "They are\nthe same"="They are the same"),
    task = factor(task, levels=c("cd", "ab"), labels=c("Aligned", "Unaligned")),
    chart = factor(chart, levels=c("Vertical", "Horizontal", "Horizontal wide")),
    zoom = factor(zoom, levels=c("no", "yes"))
  ) %>% 
  ggplot(aes(x = response3)) + 
  geom_bar(aes(fill = zoom, weight = weight), position="fill") +
  facet_grid(.~chart) + 
  scale_fill_manual("Zoom", values=c("grey", "#1274F3")) +
  xlab("") + 
  ggtitle("\nby chart") +
  theme(axis.text.x = element_text(angle=0, vjust = 0.5)) + 
#  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  scale_y_continuous("Percentage", breaks=seq(0,1, by=0.25), 
                     labels = seq(0,1, by=0.25)*100) + 
  geom_rect(
    aes(
      xmin = as.numeric(response3) - 0.45,
      ymin = zoom2 - qnorm(0.975)* `se.as.numeric(zoom)`,
      xmax = as.numeric(response3) + 0.45,
      ymax = zoom2 + qnorm(0.975)* `se.as.numeric(zoom)`,
    ), 
    fill = "black", alpha = 0.25, #colour = "white", linewidth = 0.25,
    data = confint_zoom2
  )  + 
  geom_segment(
    aes(
      x = as.numeric(response3) - 0.45,
      y = zoom2,
      xend = as.numeric(response3) + 0.45,
      yend = zoom2
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = confint_zoom2
  )

zt + zd + plot_layout(widths=c(6, 9), guides='collect')

>>>>

---
title: "What do they see? Testing perceptual accuracy in a U.S. general population survey using stacked bar charts"
short-title: Testing charts in surveys # running title
bibliography: references.bib
abstract: "The use of visuals is a key component in scientific communication. Decisions about the design of a data visualization should be informed by what design elements best support the audience's ability to perceive and understand the components of the data visualization. We build on the foundations of Cleveland and McGill's work in graphical perception, employing a large, nationally-representative, probability-based panel of survey respondents to test perception in stacked bar charts. Our findings provide actionable guidance for data visualization practitioners to employ in their work."
keywords: 'graphical perception, data visualization, survey data'
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
  \usepackage{booktabs}
  \usepackage{graphicx}
  \usepackage{hyperref}
  \usepackage{setspace}
  \usepackage{lscape}
documentclass: jds
format:
  pdf: 
    includes:
      in_header: 'gt_packages.sty'
    include-in-header: 
      text: |
        \usepackage{lscape}
        \author[1]{Kiegan Rice\thanks{Corresponding author. Email: rice-kiegan@norc.org}}
        \author[2]{Heike Hofmann\footnote{Email: hofmann@iastate.edu}}
        \author[1]{Nola du Toit\footnote{Email: dutoit-nola@norc.org}}
        \author[1]{Edward Mulrow\footnote{Email: mulrow-edward@norc.org}}
        \affil[1]{NORC at the University of Chicago}
        \affil[2]{Department of Statistics, Iowa State University}
    keep-tex: true
  html: default
editor: 
  markdown: 
    wrap: sentence
---

@@

@@

```{r}
#| label: fig-tasks
#| fig-cap: "Each of the visual stimuli presented to viewers. In each bar chart, two pieces are marked. The larger piece in each chart are pieces B and D. "
#| out-width: 90%
knitr::include_graphics("images/stimuli-overview.png")
```

@@

| Name      | Date       |      \# Participants |                Effective Sample Size |         Sum of weights $\sum_i w_i$ |                                  $\lambda_i$ |
|------------|------------|-----------:|-----------:|-----------:|-----------:|
| Round 1   | April 2022 | `r nrow(round1_adj)` | `r round(neff(round1_adj$weight),1)` | `r round(sum(round1_adj$weight),1)` |        $\lambda_1$ = `r round(lambdas[1],3)` |
| Round 2   | May 2022   | `r nrow(round2_adj)` | `r round(neff(round2_adj$weight),1)` | `r round(sum(round2_adj$weight),1)` | $\lambda_2$ = `r sprintf("%.3f",lambdas[2])` |
| Round 3   | June 2022  | `r nrow(round3_adj)` | `r round(neff(round3_adj$weight),1)` | `r round(sum(round3_adj$weight),1)` |        $\lambda_3$ = `r round(lambdas[3],3)` |
| **Total** | ---        |  `r nrow(round_123)` |  `r round(neff(round_123$weight),1)` |                                     |                                              |

@@

```{r}
#| label: fig-response-123
#| fig-cap: "Responses by task across the designs. The overlaid rectangles represent 95% confidence intervals. The letters in blue and orange encode significances between pairwise proportions: two bars have a significantly different proportion (at a 5% significance level) if they do not share a letter. There is no significant difference between the designs for wrong responses."
#| fig-height: 4
#| fig-width: 8
#| out-width: 90%

@@

where $\ell \in \{1, 2\}$; $i \in \{1, 2\}$ is the comparison type (1 = Aligned, 2 = Unaligned), and $j \in \{1, 2, 3\}$ is the chart design, with 1 = Vertical, 2 = Horizontal, and 3 = Horizontal wide.
The estimated values and 95% confidence intervals are shown in @tbl-rounds-123.
The lowercase letters indicate significances between pairs of estimates: two estimates within the same column are significantly different at 5% level if they do not have any letter in common [@piephoAlgorithmLetterBasedRepresentation2004].

@@

```{r}
#| label: fig-zoom-123
#| fig-cap: "Rates of zooming by task, chart design, and correctness. The overlaid rectangles represent 95% confidence intervals. There are significantly higher rates of zooming by respondents for the vertical task compared to both horizontal tasks; rates of zooming do not differ by correctness of response or alignment in the stimulus."
#| fig-height: 4.25
#| fig-width: 7
#| out-width: 80%

confint_zoom1 <- svyby(formula = ~zoom, by = ~task+response3, design = survey_123, svyciprop) %>%
  mutate(task = factor(task, levels=c("cd", "ab"), labels=c("Aligned", "Unaligned")), 
         zoom2 = zoom, 
        response3 = factor(response3, levels=rev(c("Correct", "They are the same", "Wrong")))) %>%
  dplyr::select(-c("zoom"))

confint_zoom2 <- svyby(formula = ~zoom, by = ~chart+response3, design = survey_123, svyciprop) %>%
  mutate(chart = factor(chart, levels=c("Vertical", "Horizontal", "Horizontal wide")), 
         zoom2 = zoom, 
        response3 = factor(response3, levels=rev(c("Correct", "They are the same", "Wrong")))) %>%
  dplyr::select(-c("zoom"))

zt <- round_123_long %>% 
  mutate(
    response3 = c("Wrong", "They are the same", "Correct")[correct+correct3+1],
    response3 = factor(response3, levels=rev(c("Correct", "They are the same", "Wrong"))),
    task = factor(task, levels=c("cd", "ab"), labels=c("Aligned", "Unaligned")),
    chart = factor(chart, levels=c("Vertical", "Horizontal", "Horizontal wide")),
    zoom = factor(zoom, levels=c("no", "yes"))
  ) %>% 
  ggplot(aes(x = response3)) + 
  geom_bar(aes(fill = zoom, weight = weight), position="fill") +
  facet_grid(.~task) + 
  theme(axis.text.x = element_text(angle=30, hjust = 1)) + 
  scale_fill_manual("Zoom", values=c("grey", "#1274F3")) +
  xlab("") + 
  ggtitle("Zooming behavior\nby task") +
  scale_y_continuous("Percentage", breaks=seq(0,1, by=0.25), 
                     labels = seq(0,1, by=0.25)*100) + 
  geom_rect(
    aes(
      xmin = as.numeric(response3) - 0.45,
      ymin = zoom2 - qnorm(0.975)* `se.as.numeric(zoom)`,
      xmax = as.numeric(response3) + 0.45,
      ymax = zoom2 + qnorm(0.975)* `se.as.numeric(zoom)`,
    ), 
    fill = "black", alpha = 0.25, #colour = "white", linewidth = 0.25,
    data = confint_zoom1
  )  + 
  geom_segment(
    aes(
      x = as.numeric(response3) - 0.45,
      y = zoom2,
      xend = as.numeric(response3) + 0.45,
      yend = zoom2
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = confint_zoom1
  )

zd <- round_123_long %>% 
  mutate(
    response3 = c("Wrong", "They are the same", "Correct")[correct+correct3+1],
    response3 = factor(response3, levels=rev(c("Correct", "They are the same", "Wrong"))),
    task = factor(task, levels=c("cd", "ab"), labels=c("Aligned", "Unaligned")),
    chart = factor(chart, levels=c("Vertical", "Horizontal", "Horizontal wide")),
    zoom = factor(zoom, levels=c("no", "yes"))
  ) %>% 
  ggplot(aes(x = response3)) + 
  geom_bar(aes(fill = zoom, weight = weight), position="fill") +
  facet_grid(.~chart) + 
  scale_fill_manual("Zoom", values=c("grey", "#1274F3")) +
  xlab("") + 
  ggtitle("\nby chart") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  scale_y_continuous("Percentage", breaks=seq(0,1, by=0.25), 
                     labels = seq(0,1, by=0.25)*100) + 
  geom_rect(
    aes(
      xmin = as.numeric(response3) - 0.45,
      ymin = zoom2 - qnorm(0.975)* `se.as.numeric(zoom)`,
      xmax = as.numeric(response3) + 0.45,
      ymax = zoom2 + qnorm(0.975)* `se.as.numeric(zoom)`,
    ), 
    fill = "black", alpha = 0.25, #colour = "white", linewidth = 0.25,
    data = confint_zoom2
  )  + 
  geom_segment(
    aes(
      x = as.numeric(response3) - 0.45,
      y = zoom2,
      xend = as.numeric(response3) + 0.45,
      yend = zoom2
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = confint_zoom2
  )

zt + zd + plot_layout(widths=c(5, 7.5), guides='collect')
